---
title: "Choosing the knobs"
subtitle: "Tuning, folds, and what cross-validation is actually for"
author: "Georgy Tarasenko · GOVT 6039 · Discussion Section 2"
date: "9 September 2026"
format:
  revealjs:
    theme: [default, custom.scss]
    embed-resources: true
    slide-number: c/t
    width: 1300
    height: 750
    fig-width: 8
    fig-height: 3.9
    fig-format: svg
    code-line-numbers: false
    highlight-style: github
    transition: none
  beamer:
    aspectratio: 169
    theme: default
    colortheme: dove
    include-in-header: preamble.tex
    filters: [spans.lua]
    slide-level: 2
    fig-width: 7.5
    fig-height: 3.4
    fig-format: pdf
    code-block-font-size: \scriptsize
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| echo: false
#| cache: false
set.seed(6039)
library(ggplot2)
ink <- "#1f2937"; acc <- "#b45309"; gry <- "#9aa1ab"
th <- theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background  = element_rect(fill = "transparent", colour = NA),
        panel.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA),
        plot.title = element_blank(),
        legend.key.height = unit(13, "pt"))
theme_set(th)
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
```

## Where we are

Lecture 3, on $k$ in $k$-NN:

> "$k$ is what is called a **tuning parameter**; in this case it governs how much to smooth our function."

Lecture 4 closed on two tasks:

> **Model selection** — choosing a model's shape or its tuning parameters.
>
> **Model assessment** — how good is the model we chose?

Today: the first task on real data, and what the second one costs when the folds are drawn carelessly.

::: {.notes}
Read both quotes out. One sentence: CV can do either job, not both with the same folds.
:::

## Parameters and hyperparameters

**Parameter.** $\hat\beta = \arg\min_b \mathrm{RSS}(b)$. The data pick it.

**Hyperparameter.** $k$. Polynomial degree. $K$, the number of folds. Tomorrow, $\lambda$.

Why can the data not pick these the same way?

. . .

Because training MSE is [monotone in flexibility]{.accent}. Minimise it over $k$ and you get $k=1$, every time — lecture 3's high-variance corner. Over the degree, the highest degree on offer.

So the choice is made on rows the model has not seen. That is the second loop.

## The knobs you already own

| knob | where you met it | what it controls |
|:--|:--|:--|
| $k$ | lecture 3 | how much to smooth |
| degree | lecture 4 | how curved $f$ may be |
| $K$ | lecture 4 | bias–variance of the CV estimate itself |
| $\lambda$ | tomorrow | how hard to shrink |
| [**the folds**]{.accent} | [**today**]{.accent} | [**what "out of sample" means**]{.accent} |

The last one is not usually called a hyperparameter. It behaves exactly like one.

# Part 1 · The knob you set without noticing

## The most common model in political science

You have a treatment $D$, a moderator $X$, and a theory that the effect of $D$ depends on $X$. So you fit

$$Y = \beta_0 + \beta_D D + \beta_X X + \beta_{DX}\,(D \times X) + \varepsilon$$

and report the marginal effect

$$\widehat{ME}(x) = \hat\beta_D + \hat\beta_{DX}\, x$$

That is [a straight line in $x$]{.accent}, imposed before you looked at anything.

Nobody calls it a tuning parameter. It is one: flexibility, fixed at its most rigid setting, by convention.

## Hainmueller, Mummolo & Xu (2019)

*Political Analysis* 27(2), "How Much Should We Trust Estimates from Multiplicative Interaction Models?"

Their argument, in our vocabulary: the linear interaction model is an untuned choice, and it fails in two ways.

1. The true $ME(x)$ may not be linear in $x$.
2. There may be almost no data in parts of the $X$ range, so the line is [extrapolating]{.accent}, not estimating.

Their remedy is one you already know: **let the flexibility be chosen**. Bin the moderator, or use a kernel estimator whose bandwidth is picked [by cross-validation]{.accent}.

## Vernby (2013), *AJPS*

Sweden extends the local franchise to noncitizens. Does school spending follow?

- 183 municipalities
- $D$ = noncitizen share of the electorate
- $X$ = tax base, the moderator
- $Y$ = change in education spending

```{r}
#| label: load-interflex
u <- paste0("https://raw.githubusercontent.com/xuyiqing/",
            "interflex/master/data/interflex.RData")
if (!file.exists("interflex.RData")) download.file(u, "interflex.RData", mode = "wb")
load("interflex.RData")
v <- app_vernby2013
dim(v)
```

## The published shape

```{r}
#| label: vernby-lm
m  <- lm(school_diff ~ noncitvotsh * Taxbase2, data = v)
round(summary(m)$coefficients, 3)
```

```{r}
#| label: vernby-me
#| echo: false
b  <- coef(m); s <- summary(m)$coefficients
qs <- quantile(v$Taxbase2, c(.05, .25, .5, .75, .95))
lin_me <- b["noncitvotsh"] + b["noncitvotsh:Taxbase2"] * qs
knitr::kable(t(setNames(round(lin_me, 1), c("p05","p25","p50","p75","p95"))),
             caption = "Implied marginal effect across the tax base")
```

A clean story: enfranchisement raises school spending in poor municipalities and [lowers]{.accent} it in rich ones — the interaction carries $t = `r fmt(s[4,3])`$.

## Now relax the knob

HMX's binning estimator: cut the moderator into thirds, estimate the effect separately inside each. No functional form across bins.

```{r}
#| label: vernby-bins
v$bin <- cut(v$Taxbase2, quantile(v$Taxbase2, c(0, 1/3, 2/3, 1)),
             labels = c("low", "mid", "high"), include.lowest = TRUE)
binest <- do.call(rbind, lapply(levels(v$bin), function(L) {
  z <- v[v$bin == L, ]; f <- lm(school_diff ~ noncitvotsh, data = z)
  data.frame(third = L, n = nrow(z), xbar = mean(z$Taxbase2),
             ME = coef(f)[2], se = summary(f)$coefficients[2, 2])
}))
knitr::kable(binest, digits = 2, row.names = FALSE)
```

The decline is there. The reversal is not.

## Now relax the knob

```{r}
#| label: plot-vernby
#| echo: false
xg <- seq(min(v$Taxbase2), max(v$Taxbase2), length.out = 200)
line <- data.frame(x = xg, me = b[2] + b[4] * xg)
bins <- transform(binest, lo = ME - 1.96 * se, hi = ME + 1.96 * se)
ggplot() +
  geom_hline(yintercept = 0, colour = gry, linewidth = .4) +
  geom_line(data = line, aes(x, me), colour = acc, linewidth = 1) +
  geom_errorbar(data = bins, aes(xbar, ymin = lo, ymax = hi), width = .02,
                colour = ink, linewidth = .6) +
  geom_point(data = bins, aes(xbar, ME), colour = ink, size = 3) +
  annotate("text", x = .53, y = b[2] + b[4] * .53 - 4, hjust = 0, vjust = 1,
           label = "linear interaction", colour = acc, size = 4.4) +
  annotate("text", x = 1.02, y = bins$hi[3], hjust = .5, vjust = -0.8,
           label = "estimated within\nthirds of the moderator",
           colour = ink, size = 4, lineheight = .95) +
  coord_cartesian(ylim = c(-13, 24)) +
  labs(x = "tax base (moderator)", y = "ME of noncitizen vote share")
```

## What actually happened

The line has to keep going. The data stop.

- $\widehat{ME}$ at the top of the range: $`r fmt(lin_me[5], 1)`$ from the line, $+`r fmt(binest$ME[3], 1)` \pm `r fmt(1.96*binest$se[3], 1)`$ from the data in that region.
- The negative number is not an estimate. It is the straight line continuing past where anything was observed.
- The $t = `r fmt(s[4,3])`$ on the interaction is real, and it is about the **slope** — not about any municipality crossing zero.

[The rigid setting did not just lose precision. It produced a sign that is not in the data.]{.accent}

## The other outcome: Huddy, Mason & Aarøe (2015), *APSR*

"Expressive Partisanship." Does partisan identity amplify anger at threats to one's party?

```{r}
#| label: hma-lm
h  <- app_hma2015
mh <- lm(totangry ~ threat * pidentity, data = h)
round(summary(mh)$coefficients, 3)
```

```{r}
#| label: hma-store
#| echo: false
bh <- coef(mh); sh <- summary(mh)$coefficients
```

$n = `r nrow(h)`$, threat randomised, $\hat\beta_{DX} = `r fmt(bh[4], 3)`$ $(`r fmt(sh[4,2], 3)`)$, $t = `r fmt(sh[4,3])`$.

Same check, with lecture 3's tool: estimate $E[Y \mid X]$ **separately in each arm** by $k$-NN, choose $k$ by 10-fold CV. The marginal effect is the gap between the two local averages.

## Choosing $k$ inside each arm

```{r}
#| label: knn1
knn1 <- function(xtr, ytr, xte, ks) {           # all k at once
  M <- outer(xte, xtr, function(a, b) (a - b)^2)
  out <- matrix(NA_real_, length(xte), length(ks))
  for (i in seq_along(xte)) {
    o <- order(M[i, ]); out[i, ] <- (cumsum(ytr[o]) / seq_along(o))[ks]
  }
  out
}
cvk1 <- function(x, y, ks, K = 10) {            # lecture 4, verbatim
  set.seed(6039); f <- sample(rep(1:K, length.out = length(y)))
  M <- matrix(NA_real_, K, length(ks))
  for (j in 1:K) M[j, ] <- colMeans((y[f == j] -
                    knn1(x[f != j], y[f != j], x[f == j], ks))^2)
  colMeans(M)
}
ks   <- c(5, 10, 25, 50, 100, 200, 400)
best <- sapply(0:1, function(d) {
  z <- h[h$threat == d, ]; ks[which.min(cvk1(z$pidentity, z$totangry, ks))]
})
setNames(best, c("control arm", "treated arm"))
```

CV asks for $k = `r best[1]`$ and $k = `r best[2]`$. [The data want heavy smoothing.]{.accent}

## One dataset, three answers

```{r}
#| label: hma-me
g  <- seq(quantile(h$pidentity, .05), quantile(h$pidentity, .95), length.out = 9)
me <- function(k) knn1(h$pidentity[h$threat == 1], h$totangry[h$threat == 1], g, k)[, 1] -
                  knn1(h$pidentity[h$threat == 0], h$totangry[h$threat == 0], g, k)[, 1]
rbind(`linear interaction` = bh["threat"] + bh["threat:pidentity"] * g,
      `k-NN, CV's k`       = me(max(best)),
      `k-NN, k = 5`        = me(5)) |> round(2)
```

```{r}
#| label: plot-hma
#| echo: false
#| fig-height: 3.2
lab <- c(sprintf("linear interaction (degree 1, untuned)"),
         sprintf("k-NN, k chosen by CV (k = %d)", max(best)),
         "k-NN, k = 5 (too flexible)")
d2 <- rbind(data.frame(x = g, me = bh["threat"] + bh["threat:pidentity"] * g, w = lab[1]),
            data.frame(x = g, me = me(max(best)), w = lab[2]),
            data.frame(x = g, me = me(5), w = lab[3]))
d2$w <- factor(d2$w, levels = lab)
ggplot(d2, aes(x, me, colour = w)) +
  geom_hline(yintercept = 0, colour = gry, linewidth = .4) +
  geom_line(linewidth = .9) + geom_point(size = 1.8) +
  scale_colour_manual(values = c(acc, ink, gry)) +
  labs(x = "partisan identity (moderator)", y = "ME of threat", colour = NULL) +
  theme(legend.position = "inside", legend.position.inside = c(.32, .86))
```

## One dataset, three answers

| setting | what you would report | |
|:--|:--|:--|
| $k = 5$, untuned | effect non-monotonic, turns [negative]{.accent} | noise |
| degree 1, untuned | effect rises steeply, 0.16 → 0.44 | overstated slope |
| $k$ from CV | effect rises gently, 0.25 → 0.37 | [what the data support]{.accent} |

Two lessons, and the second is the one people forget:

1. An untuned knob can manufacture a finding — in either direction.
2. Here, tuning largely [vindicates]{.accent} the published paper. That is a result. You only get to say it because you checked.

## Part 1, in one line

::: {.big}
[Tuning is not "make the model fancier."]{.accent}
:::

It is letting the data set the flexibility — and being willing to hear the answer "stay simple."

HMX ran this check on **46 interaction effects from 22 articles** in the *APSR*, *AJPS*, *CPS*, *IO* and *JOP*, published 2006–2015.

| | share of cases |
|:--|--:|
| the effect changes linearly in the moderator | [48%]{.accent} |
| cannot reject that the effect is equal at typical low and high moderator values | [70%]{.accent} |

## And the knob is still being argued about

This is not settled methodology from a textbook. It is a live fight.

- **Simonsohn (2024)**, *AMPPS*, and the companion "Dear Political Scientists: Don't Bin, GAM Instead": binning is the wrong fix; model the curve with a GAM.
- **Hainmueller, Liu, Liu, Mummolo & Xu (2025)**: the critique benchmarks against the wrong estimand; the kernel estimator does recover the conditional marginal effect; GAM is for exploration, not for this.

Both sides agree on what matters today: [the flexibility is a choice, it has to be made explicitly, and the data have to be allowed to inform it.]{.accent}

They disagree about *how*. That disagreement is open, and it is open to you.

# Part 2 · The folds are a knob too

## A question you can actually ask

How much of a country's level of democracy is accounted for by its structural conditions?

```{r}
#| label: vdem-prep
uv <- paste0("https://raw.githubusercontent.com/vdeminstitute/",
             "vdemdata/master/data/vdem.RData")
if (!file.exists("vdem_slim.rds")) {
  if (!file.exists("vdem.RData")) download.file(uv, "vdem.RData", mode = "wb")
  load("vdem.RData")
  saveRDS(vdem[, c("country_name","year","v2x_polyarchy","e_gdppc","e_pop",
                   "e_total_resources_income_pc","e_regionpol_6C")], "vdem_slim.rds")
  rm(vdem); gc()
}
vd <- readRDS("vdem_slim.rds")
vd <- vd[vd$year >= 1960 & vd$year <= 2023, ]; vd <- vd[complete.cases(vd), ]
vd$lgdp <- log(pmax(vd$e_gdppc, .01)); vd$lpop <- log(pmax(vd$e_pop, .01))
vd$lres <- log1p(pmax(vd$e_total_resources_income_pc, 0))
regmat <- model.matrix(~ factor(e_regionpol_6C) - 1, data = vd)[, -1, drop = FALSE]
X <- cbind(as.matrix(vd[, c("lgdp","lpop","lres","year")]), regmat)
y <- vd$v2x_polyarchy; ctry <- vd$country_name
c(n = nrow(vd), countries = length(unique(ctry)), var_y = round(var(y), 4))
```

V-Dem v15, `r min(vd$year)`–`r max(vd$year)`. $Y$ is the electoral democracy index; $X$ is log GDP per capita, log population, resource income, region and year.

## $k$-NN and 10-fold CV, exactly as on Tuesday

```{r}
#| label: vdem-helpers
knn_all <- function(Xtr, ytr, Xte, ks) {
  mu <- colMeans(Xtr); sdv <- apply(Xtr, 2, sd); sdv[sdv == 0] <- 1
  A <- scale(Xtr, mu, sdv); B <- scale(Xte, mu, sdv)   # standardise INSIDE the fold
  Dm <- outer(rowSums(B^2), rowSums(A^2), "+") - 2 * B %*% t(A)
  out <- matrix(NA_real_, nrow(B), length(ks))
  for (i in seq_len(nrow(B))) {
    o <- order(Dm[i, ]); out[i, ] <- (cumsum(ytr[o]) / seq_along(o))[ks]
  }
  out
}
cv <- function(fold, ks) {
  M <- matrix(NA_real_, max(fold), length(ks))
  for (j in 1:max(fold)) {
    tr <- fold != j; te <- fold == j
    M[j, ] <- colMeans((y[te] - knn_all(X[tr, ], y[tr], X[te, ], ks))^2)
  }
  colMeans(M)
}
ks2 <- c(1, 2, 3, 5, 10, 25, 50, 100, 250, 500)
```

```{r}
#| label: vdem-cv
#| cache: true
set.seed(6039); fr <- sample(rep(1:10, length.out = nrow(vd)))          # random folds
set.seed(6039); uc <- unique(ctry)
fb <- sample(rep(1:10, length.out = length(uc)))[match(ctry, uc)]       # whole countries
A <- cv(fr, ks2); B <- cv(fb, ks2)
```

## Modernization theory, solved

```{r}
#| label: vdem-random
#| echo: false
knitr::kable(data.frame(k = ks2, `CV MSE` = round(A, 4),
                        `implied R2` = round(1 - A / var(y), 3), check.names = FALSE),
             row.names = FALSE)
```

$k = `r ks2[which.min(A)]`$ wins, at MSE `r fmt(min(A), 4)` — an implied $R^2$ of [`r fmt(1 - min(A)/var(y), 2)`]{.accent}.

Five structural variables and a local average of two neighbours account for 98% of the variation in democracy across the postwar world.

::: {.notes}
STOP HERE. Ask the room whether they believe it. Let them sit with it.
:::

## Do we believe this?

::: {.big}
[Five variables. Two neighbours. $R^2 = `r fmt(1 - min(A)/var(y), 2)`$.]{.accent}
:::

. . .

The folds were drawn at random over **country-years**.

Denmark 1994 and Denmark 1995 are nearly the same row. A random fold puts one in training and the other in the held-out set.

. . .

So hold out whole **countries** instead.

## Draw the folds by country instead

```{r}
#| label: vdem-blocked
#| echo: false
kr <- ks2[which.min(A)]
knitr::kable(data.frame(
  ` ` = c("scored on held-out country-years", "scored on held-out **countries**"),
  `CV MSE` = c(round(min(A), 4), round(B[which.min(A)], 4)),
  `implied R2` = round(c(1 - min(A)/var(y), 1 - B[which.min(A)]/var(y)), 3),
  check.names = FALSE), row.names = FALSE)
```

The same $k = `r kr`$ model. [**`r round(B[which.min(A)]/min(A))` times worse.**]{.accent}

Nothing changed but which rows went into which fold.

## The whole picture

```{r}
#| label: plot-vdem
#| echo: false
#| fig-height: 3.4
ob <- mean(sapply(1:10, function(j) {
  df <- as.data.frame(X); tr <- fb != j; te <- fb == j
  mean((y[te] - predict(lm(y[tr] ~ ., data = df[tr, ]), df[te, ]))^2)
}))
d3 <- rbind(data.frame(k = ks2, mse = A, how = "random 10-fold"),
            data.frame(k = ks2, mse = B, how = "blocked by country"))
d3$how <- factor(d3$how, levels = c("random 10-fold", "blocked by country"))
pick <- data.frame(k = c(ks2[which.min(A)], ks2[which.min(B)]),
                   mse = c(min(A), min(B)),
                   how = factor(c("random 10-fold", "blocked by country"),
                                levels = levels(d3$how)))
ggplot(d3, aes(k, mse, colour = how)) +
  geom_hline(yintercept = var(y), linetype = "22", colour = gry) +
  annotate("text", x = 1, y = var(y), hjust = 0, vjust = 1.5, size = 4,
           colour = "grey40", label = "predicting the global mean") +
  geom_hline(yintercept = ob, linetype = "42", colour = ink, linewidth = .4) +
  annotate("text", x = 500, y = ob, hjust = 1, vjust = -0.7, size = 4,
           colour = ink, label = "OLS, blocked") +
  geom_line(linewidth = .9) + geom_point(size = 2) +
  geom_point(data = pick, aes(k, mse), size = 5.5, shape = 21, stroke = 1.1, fill = NA) +
  scale_x_log10(breaks = ks2) + scale_colour_manual(values = c(acc, ink)) +
  labs(x = "k (log scale)", y = "CV MSE", colour = NULL) +
  theme(legend.position = "bottom")
```

Circles mark the $k$ each procedure selects.

## Two things moved, not one

```{r}
#| label: vdem-summary
#| echo: false
oa <- mean(sapply(1:10, function(j) {
  df <- as.data.frame(X); tr <- fr != j; te <- fr == j
  mean((y[te] - predict(lm(y[tr] ~ ., data = df[tr, ]), df[te, ]))^2)
}))
knitr::kable(data.frame(
  ` ` = c("the k that wins", "its CV MSE", "OLS"),
  `random folds` = c(as.character(ks2[which.min(A)]), fmt(min(A), 4), fmt(oa, 4)),
  `blocked by country` = c(as.character(ks2[which.min(B)]), fmt(min(B), 4), fmt(ob, 4)),
  check.names = FALSE), row.names = FALSE, align = "lrr")
```

1. The **error estimate** was inflated by a factor of `r round(B[which.min(A)]/min(A))`.
2. The **hyperparameter** the procedure selected moved from $k = `r ks2[which.min(A)]`$ to $k = `r ks2[which.min(B)]`$.
3. Under honest folds, $k$-NN at its best (`r fmt(min(B), 4)`) barely edges plain OLS (`r fmt(ob, 4)`). The flexible model buys almost nothing.

If you tune with leaky folds you do not merely overstate your error. [You choose the wrong model.]{.accent}

## Why $k=2$ wins on random folds

The two nearest neighbours of Denmark 1995 are Denmark 1994 and Denmark 1996.

$k = 2$ is not a model of democratisation. It is a [lookup table keyed on country identity]{.accent}, and random folds hand it the key.

Block by country and the key is gone, so the estimator has to fall back on structure — which is worth an $R^2$ of about `r fmt(1 - min(B)/var(y), 2)`, not `r fmt(1 - min(A)/var(y), 2)`.

## The rule, restated

Tuesday's version:

> The fold must redo every data-dependent step: screening, standardising, imputing, choosing $k$.

Today's addition:

> [And the folds must respect the dependence in your data.]{.accent}

Random $K$-fold assumes the rows are exchangeable draws. Ask whether yours are before you accept the number.

## Where this bites in your own work

| your data | the unit to hold out |
|:--|:--|
| country-year or state-year panel | country / state |
| survey with respondents seen twice | respondent |
| students in schools, voters in precincts | school / precinct |
| districts in space | a spatial block |
| dyads | both members of the dyad |
| anything with a time order and a forecast | the future |

If two rows would never be separated in the real prediction problem, they should not be separated by your folds.

# What to take away

## Selection and assessment, one more time

The number that chose your model cannot also be the number that grades it.

- Selection: an inner loop, over the grid, on training rows only.
- Assessment: an outer loop, or a test set you touch once.
- Do both with one set of folds and the reported error is the best of many draws.

This is Tuesday's slide 19 and today's Part 2. Same disease, two organs.

## What to report

1. The **grid** you searched, and how many fits it took.
2. The **metric**, and why it suits the outcome.
3. **How the folds were drawn**, and what unit was held out.
4. Whether the reported error came from the **same loop** that chose the model.
5. For any interaction: the **support** across the moderator, and one estimate that does not assume linearity.

None of this is expensive. All of it is usually missing.

## Three sentences you could have written today

1. *"Noncitizen enfranchisement reduces school spending in wealthy municipalities."* — an artifact of the straight line, at a tax base where no municipality was observed.

2. *"Structural conditions account for `r fmt(100*(1 - min(A)/var(y)), 0)`% of the variation in democracy."* — an artifact of the folds, which handed the model the country's own past.

3. *"The effect declines in the tax base, and flexible estimators do not beat OLS on V-Dem."* — what survives.

The first two are more interesting. That is the problem.

## For tomorrow and for the problem set

- Tomorrow is penalisation: $\lambda$ is the same object as $k$, and is chosen the same way. Everything today transfers.
- On PS1: when you fit anything with a knob, show the grid and say how the folds were drawn.
- If your data have repeated units, block on them and report both numbers. The gap is itself a finding.

::: {.big}
[Questions?]{.accent}
:::

## References {.allowframebreaks}

::: {.smaller}

Hainmueller, J., J. Mummolo & Y. Xu. 2019. "How Much Should We Trust Estimates from Multiplicative Interaction Models? Simple Tools to Improve Empirical Practice." *Political Analysis* 27(2): 163–192.

Vernby, K. 2013. "Inclusion and Public Policy: Evidence from Sweden's Introduction of Noncitizen Suffrage." *AJPS* 57(1): 15–29.

Huddy, L., L. Mason & L. Aarøe. 2015. "Expressive Partisanship: Campaign Involvement, Political Emotion, and Partisan Identity." *APSR* 109(1): 1–17.

Simonsohn, U. 2024. "Interacting with Curves: How to Validly Test and Probe Interactions in the Real (Nonlinear) World." *AMPPS* 7(1).

Hainmueller, J., J. Liu, Z. Liu, J. Mummolo & Y. Xu. 2025. "A Response to Recent Critiques of Hainmueller, Mummolo and Xu (2019)." arXiv:2502.05717.

Coppedge, M. et al. 2025. *V-Dem Country-Year Dataset v15*. Varieties of Democracy Institute.

Hastie, T., R. Tibshirani & J. Friedman. 2009. *The Elements of Statistical Learning*, 2nd ed., §7.10.2.

Roberts, D. R. et al. 2017. "Cross-validation strategies for data with temporal, spatial, hierarchical, or phylogenetic structure." *Ecography* 40(8): 913–929.

Application data via the `interflex` package (Hainmueller, Mummolo, Xu & Liu).
:::
